---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# wisclabmisc

<!-- badges: start -->
<!-- badges: end -->

The goal of wisclabmisc is to reuse analysis functions across projects.

## Installation

You can install the development version of wisclabmisc from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("tjmahr/wisclabmisc")
```
## gamlss helpers

`mem_gamlss()` (memory gamlss) provides a drop-in replacement for the `gamlss()`
function.

```{r example, message = FALSE}
library(wisclabmisc)
library(gamlss)
library(tidyverse)

data <- as.data.frame(nlme::Orthodont)
model <- mem_gamlss(distance ~ age, data = data)
```

The only difference between `mem_gamlss()` and `gamlss()` is that the modified
version includes a bundle of data in `.user` that records the original dataset,
session information and the call used to fit the model. 

```{r}
str(model$.user, max.level = 1)
```

gamlss does not store the data as part of the model object, and we need the
dataset because prediction and centile prediction often fails without the
dataset:

```{r, error = TRUE}
newdata <- distinct(data, age)
centiles.pred(
  model, 
  cent = c(25, 50, 75),
  xname = "age", 
  xvalues = newdata$age, 
  plot = FALSE
)
```

But including the original dataset works:

```{r, error = TRUE}
centiles.pred(
  model, 
  cent = c(25, 50, 75),
  xname = "age", 
  xvalues = newdata$age, 
  plot = FALSE,
  data = model$.user$data
)
```

This package provides `predict_centiles()` as a streamlined version of the
above code, but:

- assumes the model was fitted with `mem_gamlss()`
- returns a tibble
- keeps the predictor name (here, `age` instead of `x`)
- prefixes the centiles with `q` (for quantile)

```{r}
centiles <- predict_centiles(
  newdata,
  model, 
  cent = c(25, 50, 75)
)
centiles
```

Those predicted centiles are in wide format. We can tidy them into a long format
with `pivot_centiles_longer()`. This also includes `.pair` column that helps
mark commonly paired quantiles 25:75, 10:90, and 5:95. 

```{r}
pivot_centiles_longer(centiles)
```

