---
title: "A Workflow for brms"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
dir.create("models", showWarnings = FALSE)
library(brms)
```

```{r setup}
library(brms)
```

[brms](https://paulbuerkner.com/brms/) is the premier Bayesian regression package in R, and we use it in many of 
our projects. In this article, I describe the workflow for using brms that I have 
<strike>stumbled upon</strike> iterated over our last few analysis projects.

## Context: A set of models

Suppose that we have a dataset of age and intelligibility.

```{r}
data <- wisclabmisc::data_fake_intelligibility
data
```

We want to fit a *set of related models*. For example, we want to
compare models with different kinds of age trends using splines:

```{r}
library(brms)
library(splines)
f1 <- bf(
  intelligibility ~ age_months, 
  phi ~ age_months, 
  family = Beta
)
f2 <- bf(
  intelligibility ~ ns(age_months, 2), 
  phi ~ ns(age_months, 2),
  family = Beta
)
f3 <- bf(
  intelligibility ~ ns(age_months, 2), 
  phi ~ ns(age_months, 2),
  family = Beta
)
```

We could then write a `brm()` for each of these formulas, but that's not a good
plan. More models could arise as we start to analyze and critique the results.
We might need to set `adapt_delta` to a higher value for all the models. Aside
from the formulas, and possibly the priors, everything about the
model-fitting should be the same from model to model.

A solution for this problem to 
**make a function for fitting models from the set**. A function provides two wins:

1.  We can standardize the model fitting across models.
2.  We can indicate the similarity of the model through the organization
    of the code. A function tell us "this is where we define possible age models".

Here is what the workflow looks like at a very high level. We have a
helper function for retrieving priors, a helper function for the
default/shared arguments to `brm()`, a helper function for computing LOO
scores, and the main function for fitting models from the set.

```{r, eval = FALSE}
lookup_prior <- function(prior_slug) {
  # ...
} 

brm_args <- wisclabmisc::brms_args_create()

fit_age_model <- function(
    data,
    model_slug,
    prior_slug = "default",
    seed = NA,
    use_reloo = FALSE,
    ...
) {
  # 1. [looking up priors]

  # 2. [mapping from model_slug to formula]
  
  # 3. [creating a filename]
  
  # 4. [fitting model]
  
  # 5. [adding LOO score]
}
```

## Priors

We look up priors by using a helper function. Right now, we don't have
any different set of priors we want to compare, so there are just two
options: use some weak priors for the logit-scale mu and the log-scale
phi, or use the brms defaults.

```{r}
lookup_prior <- function(prior_slug) {
  l <- list(
    default = c(
      prior(normal(0, 1), class = b),
      prior(normal(0, 2), class = Intercept),
      prior(normal(0, 2), class = Intercept, dpar = phi),
      prior(normal(0, 2), class = b, dpar = phi)
    ),
    brms_default = NULL
  )
  l[[prior_slug]]
}
```

## Model-fitting defaults

We usually want knobs and dials of model fitting---number of chains,
number of iterations, Stan backend---to be the same from model-to-model.
So we create a function called `brm_args()` the prints out a list of
default arguments to `brm()`. We create this function using 
`wisclabmisc::brms_args_create()`.

Here are the package-provided defaults. (These defaults are just my preferred 
settings.) 

```{r}
brm_args <- wisclabmisc::brms_args_create()
brm_args()
```

But we can generate a new `brm_args()` function with different defaults. 

```{r}
brm_args <- wisclabmisc::brms_args_create(iter = 5000, chains = 1)
brm_args()
```

These `brm_args()` function lets us set default `brm()` argument values. 
Importantly, we can deviate from these defaults if we need to. Here we can
overwrite the number of chains and set a value for `adapt_delta`.

```{r}
brm_args(chains = 2, adapt_delta = .98)
```

## Another helper function for LOO computation

I also like to compute the LOO score by default. This operation can be 
time-consuming but it fortunately gets cached in the model `file`. The 
`reloo` argument gets forwarded to `brms::loo()`. If `use_reloo = TRUE`, brms 
refits the model to compute exact leave-one-out scores for problematic 
observations. 

```{r}
add_loo_criterion <- function(x, ..., use_reloo = FALSE) {
  if (use_reloo) {
    brms::add_criterion(
      x,
      criterion = "loo",
      reloo = TRUE,
      recompile = FALSE,
      ...
    )
  } else {
    brms::add_criterion(
      x,
      criterion = "loo",
      ...
    )
  }
}
```


## Model fitting function

Now, we can fit a model. Each of the steps in the function are pretty simple. 
We are generating a call to `brm()` but have to look up the prior and formula 
to use, and create a filename for the `file` argument. 

```{r}
brm_args <- wisclabmisc::brms_args_create()

fit_age_model <- function(
    data,
    model_slug,
    prior_slug = "default",
    seed = NA,
    use_reloo = FALSE, 
    ...
) {
  # 1. [looking up priors]
  prior <- lookup_prior(prior_slug)

  # 2. [mapping from model_slug to formula]
  ns <- splines::ns
  
  formulas <- list(
    linear = bf(
      intelligibility ~ age_months, 
      phi ~ age_months, 
      family = Beta
    ),
    spline_2df = bf(
      intelligibility ~ ns(age_months, 2), 
      phi ~ ns(age_months, 2),
      family = Beta
    ),
    spline_3df = bf(
      intelligibility ~ ns(age_months, 3), 
      phi ~ ns(age_months, 3),
      family = Beta
    )
  )
  
  formula <- formulas[[model_slug]]
  
  # 3. [creating a filename]
  loo_slug <- ifelse(use_reloo, "_reloo", "")
  file <- file.path(
    "models", paste0(model_slug, "_", prior_slug, loo_slug)
  )
  
  # 4. [fitting model]
  args <- brm_args(
    formula = formula,
    data = data,
    prior = prior,
    file = file,
    seed = seed,
    ...
  )
  model <- do.call(brm, args)

  # 5. [adding LOO score]
  add_loo_criterion(model, use_reloo)
}
```

Now let's fit the models and compare them.

```{r, include = FALSE}
m1 <- fit_age_model(data, "linear", seed = 20241023)
m2 <- fit_age_model(data, "spline_2df", seed = 20241023)
m3 <- fit_age_model(data, "spline_3df", seed = 20241023)
```

```{r}
m1 <- fit_age_model(data, "linear", seed = 20241023)
m2 <- fit_age_model(data, "spline_2df", seed = 20241023)
m3 <- fit_age_model(data, "spline_3df", seed = 20241023)

loo_compare(m1, m2, m3) |> 
  print(simplify = FALSE)
```


## The full workflow

Here is the full workflow from above, for easy copy-pasting as boilerplate 
for new projects. Putting this code chunk in a easy to grab place is my 
motivation for writing this article. 🤓

```{r}
lookup_prior <- function(prior_slug) {
  l <- list(
    default = c(
      prior(normal(0, 1), class = b),
      prior(normal(0, 2), class = Intercept),
      prior(normal(0, 2), class = Intercept, dpar = phi),
      prior(normal(0, 2), class = b, dpar = phi)
    ),
    brms_default = NULL
  )
  l[[prior_slug]]
}

add_loo_criterion <- function(x, ..., use_reloo = FALSE) {
  if (use_reloo) {
    brms::add_criterion(
      x,
      criterion = "loo",
      reloo = TRUE,
      recompile = FALSE,
      ...
    )
  } else {
    brms::add_criterion(
      x,
      criterion = "loo",
      ...
    )
  }
}

brm_args <- wisclabmisc::brms_args_create()

fit_age_model <- function(
    data,
    model_slug,
    prior_slug = "default",
    seed = NA,
    use_reloo = FALSE, 
    ...
) {
  # 1. [looking up priors]
  prior <- lookup_prior(prior_slug)

  # 2. [mapping from model_slug to formula]
  ns <- splines::ns
  
  formulas <- list(
    linear = bf(
      intelligibility ~ age_months, 
      phi ~ age_months, 
      family = Beta
    ),
    spline_2df = bf(
      intelligibility ~ ns(age_months, 2), 
      phi ~ ns(age_months, 2),
      family = Beta
    ),
    spline_3df = bf(
      intelligibility ~ ns(age_months, 3), 
      phi ~ ns(age_months, 3),
      family = Beta
    )
  )
  
  formula <- formulas[[model_slug]]
  
  # 3. [creating a filename]
  loo_slug <- ifelse(use_reloo, "_reloo", "")
  file <- file.path(
    "models", paste0(model_slug, "_", prior_slug, loo_slug)
  )
  
  # 4. [fitting model]
  args <- brm_args(
    formula = formula,
    data = data,
    prior = prior,
    file = file,
    seed = seed,
    ...
  )
  model <- do.call(brm, args)

  # 5. [adding LOO score]
  add_loo_criterion(model, use_reloo)
}
```

