---
title: "Utterance length imputation and weighting"
author: Tristan Mahr
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(tidyverse)
```



```{r setup}
library(wisclabmisc)
library(tidyverse)
```

In this article, I describe the background of the
`impute_values_by_length()` and `weight_lengths_with_ordinal_model()`
functions in our work and demonstrate their use. Each of these functions
had been used in different projects (for example, intelligibility and
then later articulation rate), and I had been copying and tweaking the
code from project to project. I have decided to give them a more static,
public implementation in this package. For each function, I will first
show the most recent package implementation and demonstrate the package
version, verifying that produce similar results.

For the demonstrations that follow, I used the
`data_example_intelligibility_by_length` dataset which consists of
model-simulated intelligibility scores. In short, I fit a model for
children's intelligibility scores and then asked the most to estimate
back the original data, provide a statistically faithful but jittered
version of the original dataset.

## Background

In Hustad and colleagues (2020), we modeled intelligibility data in
young children's speech. Children would hear an utterance and then 
would repeat it. The utterances started at 2 words in length, then
increased to 3 words in length, and so on in batches of 10 sentences,
all the way to 7 words in length. There was a problem, however: Not all
of the children could produce utterances at every length. Specifically,
if a child could not reliably produced 5 utterances of a given length
length, the task was halted. So given the nature of the task, if a child
had produced 5-word utterances, they also produced 2--4-word utterances
as well.

Here, in our modeled/re-simulated version of the dataset, observe how
the number of children per utterance length decreases:

```{r}
library(tidyverse)

data_example_intelligibility_by_length |> 
  count(tocs_level)
```

The length of the utterance plausibly influenced the outcome variable:
Longer utterances have more words that might help a listener understand
the sentence, for example. Therefore, it did not seem appropriate to
ignore the missing values. We used the following two-step procedure (see
the [Supplemental Materials](https://doi.org/10.23641/asha.12330956.v1)
for more detail):


1.  impute the missing values at each utterance length using values from
    shorter lengths, and do the imputation in stages, so that an
    imputed value for length *L* can be used a predictor for *L* + 1.

2.  weight each utterance length by the probability of it being produced
    for a child at a given age and take the weighted average of the
    outcome variable across all length.

Our goal for the data preparation was to produce a single-number
intelligibility score, which this final weighted average provides such a
number. With this procedure, missing data is not ignored and implausible
data (like the longest utterances at the youngest ages) is downweighted.

## Imputation

The flow for the code consists of the following steps:

  - reshape the data into a wide format,
  - train a series of linear models trained on the *observed* data,
  - predict the responses for missing values
  - reshape back to a long format

Here is the most recent non-package version of the function:

```{r}
impute_values <- function(data, var, data_train = NULL) {
  spec <- build_wider_spec_for_imputation(data, {{ var }})
  data_wide <- pivot_wider_for_imputation(data, spec)

  if (is.null(data_train)) {
    data_train <- data
  }

  data_wide_train <- pivot_wider_for_imputation(data_train, spec)
  models <- fit_imputation_models(data_wide_train)

  data_imputed <- data_wide |>
    mutate(
      y_3 = ifelse(is.na(y_3), predict(models$m_3, pick(everything())), y_3)
    ) |>
    mutate(
      y_4 = ifelse(is.na(y_4), predict(models$m_4, pick(everything())), y_4)
    ) |>
    mutate(
      y_5 = ifelse(is.na(y_5), predict(models$m_5, pick(everything())), y_5)
    ) |>
    mutate(
      y_6 = ifelse(is.na(y_6), predict(models$m_6, pick(everything())), y_6)
    ) |>
    mutate(
      y_7 = ifelse(is.na(y_7), predict(models$m_7, pick(everything())), y_7)
    )

  # handle these separately because `length_longest` could have been imputed if
  # there was a different strategy
  data_child_ll <- data_wide |>
    distinct(group, subject_num, visit_id, age_months, length_longest)

  data_original_values <- data_wide |>
    tidyr::pivot_longer_spec(spec) |>
    distinct(group, subject_num, visit_id, tocs_level, {{ var }})

  d <- data_imputed |>
    tidyr::pivot_longer_spec(spec) |>
    rename("imputed_{{ var }}" := {{ var }}) |>
    select(-length_longest) |>
    left_join(
      data_child_ll,
      by = c("group", "subject_num", "visit_id", "age_months")
    ) |>
    left_join(
      data_original_values,
      by = c("group", "subject_num", "visit_id", "tocs_level")
    ) |>
    mutate(
      imputed = ifelse(is.na({{ var }}), "imputed", "observed"),
      facet_lab = paste0(tocs_level, " words")
    )

  d
}

fit_imputation_models <- function(data) {
  list(
    m_7 = lm(y_7 ~ y_1 + y_2 + y_3 + y_4 + y_5 + y_6, data),
    m_6 = lm(y_6 ~ y_1 + y_2 + y_3 + y_4 + y_5 + length_longest, data),
    m_5 = lm(y_5 ~ y_1 + y_2 + y_3 + y_4 + length_longest, data),
    m_4 = lm(y_4 ~ y_1 + y_2 + y_3 + length_longest, data),
    m_3 = lm(y_3 ~ y_1 + y_2 + length_longest, data)
  )
}

build_wider_spec_for_imputation <- function(data, var) {
  tidyr::build_wider_spec(
    data,
    names_from = tocs_level,
    names_prefix = "y_",
    values_from = {{ var }}
  )
}

pivot_wider_for_imputation <- function(data, spec) {
  data |>
    tidyr::pivot_wider_spec(
      spec,
      id_cols = c(
        group, subject_num, visit_id, length_longest, age_months
      )
    )
}
```

This implementation is unsatisfactory for a number of reasons. It
hardcodes many unneeded variables, so to apply it to the example
dataset, I have to add filler values for those variables. It also
hardcodes the number of models and the predictors for the imputation
model. The `y_1` in the `lm()` indicates that the scores from the
single-word trials are being used for the imputation. This inclusion may
or may not be appropriate. In speaking rate studies, it doesn't make
sense to use include data from the single-word trials.

But, these caveats aside, it works:

```{r}
sim1 <- data_example_intelligibility_by_length |>
  mutate(group = "fake", visit_id = 1) |>
  filter(length_longest != 1) |> 
  rename(subject_num = child) |> 
  impute_values(sim_intelligibility)
```

And here is how we can recreate Figure 3 from the supplemental
materials:

```{r, fig.width = 6, fig.height = 5, fig.cap = "Results of imputing multiword intelligibility using length-of-longest utterance and the average intelligibilities of shorter utterance lengths."}
plotting_constants <- list(
  pal = c(
    imputed = "#C7A76C", 
    observed = "#7DB0DD", 
    "mean ± SE" = "grey30"
  ),
  guides_pal = guides(
    shape = "none",
    color = guide_legend(
      title = NULL,
      override.aes = list(
        alpha = 1,
        shape = c(16, 17, 16),
        linetype = c("blank", "blank", "solid")
      )
    )
  )
) 

plotting_constants$scale_pal <- scale_color_manual(
  values = plotting_constants$pal, 
  limits = names(plotting_constants$pal)
)

set.seed(100)
ggplot(sim1 |> filter(tocs_level != 1)) + 
  aes(
    x = length_longest, 
    y = imputed_sim_intelligibility, 
    color = imputed, 
    shape = imputed
  ) + 
  geom_jitter(width = .3, alpha = .5, height = 0) +
  stat_summary(
    aes(group = length_longest, color = "mean ± SE"), 
    fun.data = mean_se
  ) + 
  plotting_constants$scale_pal + 
  plotting_constants$guides_pal +
  facet_wrap("facet_lab") + 
  scale_y_continuous(
    "Intelligibility", 
    labels = scales::percent_format(1)
  ) +
  labs(
    x = "Length of longest utterance (observed)"
  ) + 
  theme(legend.position = "bottom", legend.justification = "right")
```

In comparison, here is the package version of this procedure. We have to
specify all of the relevant variables ahead of time, but the number of
models or the variables involved are no longer hard-coded.


```{r, fig.width = 6, fig.height = 5}
sim2 <- data_example_intelligibility_by_length |>
  filter(length_longest != 1) |> 
  impute_values_by_length(
    var_y = sim_intelligibility,
    var_length = tocs_level,
    id_cols = c(child, age_months, length_longest), 
    include_max_length = TRUE
  )

all.equal(
  sim1$imputed_sim_intelligibility, 
  sim2$sim_intelligibility_imputed
)
```

In the original models, the length of the longest utterance is used as a
continuous predictor in the imputations. In the package version this behavior is
optional and disabled by default. It does not affect the results noticeably.

```{r, fig.width = 6, fig.height = 5}
sim3 <- data_example_intelligibility_by_length |>
  filter(length_longest != 1) |> 
  impute_values_by_length(
    var_y = sim_intelligibility,
    var_length = tocs_level,
    id_cols = c(child, age_months, length_longest), 
    # changed line
    include_max_length = FALSE
  )

all.equal(
  sim2$sim_intelligibility_imputed, 
  sim3$sim_intelligibility_imputed
)

cor(
  sim2$sim_intelligibility_imputed, 
  sim3$sim_intelligibility_imputed
)

ggplot(sim2 |> filter(tocs_level != 1)) + 
  aes(
    x = length_longest, 
    y = sim_intelligibility_imputed, 
    color = sim_intelligibility_imputation, 
    shape = sim_intelligibility_imputation
  ) + 
  geom_jitter(width = .3, alpha = .5, height = 0) +
  stat_summary(
    aes(group = length_longest, color = "mean ± SE"), 
    fun.data = mean_se
  ) + 
  plotting_constants$scale_pal + 
  plotting_constants$guides_pal +
  facet_wrap("tocs_level") + 
  scale_y_continuous(
    "Intelligibility", 
    labels = scales::percent_format(1)
  ) +
  labs(
    x = "Length of longest utterance (observed)"
  ) + 
  theme(legend.position = "bottom", legend.justification = "right")
```

## Utterance weighting

Below is the most recent non-package version of the utterance length
weighting code. The basic steps are

  - preparing a dataset
  - fitting an ordinal regression model on the length of longest
    utterance using age as a predictor
  - computing the probabilities for each utterance length as a function
    of age

Originally, we had used `MASS::polr()` for the models but I think
`ordinal::clm()` is easier to use.

```{r}
prepare_longest_length_model_data <- function(data) {
  data |>
    group_by(child) |>
    mutate(
      length_longest = max(tocs_level)
    ) |>
    ungroup() |>
    distinct(child, tocs_level, age_months, length_longest) |>
    filter(tocs_level == length_longest) |>
    mutate(
      lol = factor(length_longest),
      lol_ord = ordered(lol)
    )
}

fit_longest_length_model <- function(data, df = 2) {
  ordinal::clm(
    lol_ord ~ splines::ns(age_months, df = df),
    data = data
  )
}

compute_longest_length_weights <- function(data, model) {
  d_ages <- data |>
    distinct(age_months) |>
    mutate(
      # predicted longest length by age
      lol_predicted = model |>
        predict(newdata = pick(everything()), type = "class") |>
        getElement("fit")
    )

  d_ages |>
    predict(model, newdata = _, type = "prob") |>
    getElement("fit") |>
    bind_cols(d_ages) |>
    tidyr::pivot_longer(
      cols = -c(age_months, lol_predicted),
      names_to = "lol",
      values_to = "lol_prob"
    ) |>
    group_by(age_months) |>
    mutate(
      lol_num = as.numeric(as.character(lol)),
      lol_weighted = lol_prob * lol_num
    ) |>
    arrange(-lol_num) |>
    mutate(
      prob_reach_length = cumsum(lol_prob),
      normalized_prob_reach_length = prob_reach_length / sum(prob_reach_length)
    ) |>
    ungroup()
}
```

Here we can apply this non-package implementation to the example data:

```{r, fig.width = 6, fig.height = 3}
data_lol <- data_example_intelligibility_by_length |> 
  filter(tocs_level != 1) |> 
  prepare_longest_length_model_data()

m <- fit_longest_length_model(data_lol, df = 2)

data_lol_weights <- data_lol |> 
  compute_longest_length_weights(model = m) |>
  distinct(
    age_months, 
    tocs_level = lol_num, 
    prob_reach_length, 
    normalized_prob_reach_length
  )

# data_weighted <- data_example_intelligibility_by_length |> 
#   left_join(
#     data_lol_weights,
#     by = join_by(age_months, tocs_level)
#   )
```

And here is how we can recreate the Figure from the supplemental
materials:

```{r, fig.width = 3, fig.height = 3, out.width = "40%", fig.show = "hold"}
p1 <- ggplot(data_lol_weights) + 
  aes(x = age_months, y = prob_reach_length  ) + 
  geom_line(aes(color = ordered(tocs_level))) +
  ggrepel::geom_label_repel(
    aes(label = tocs_level),
    data = data_lol_weights |> 
      filter(age_months == 30), 
    xlim = c(28, 30), 
    seed = 1
  ) +
  ggrepel::geom_label_repel(
    aes(label = tocs_level),
    data = data_lol_weights |> 
      filter(age_months == 47), 
    xlim = c(47, 49), 
    seed = 1
  ) +
  scale_color_ordinal(end = .85) +
  guides(color = "none") + 
  labs(x = "Age [months]", y = "Prob. of reaching length") +
  scale_x_continuous(limits = c(28, 49))
p2 <- p1 + 
  aes(y = normalized_prob_reach_length) + 
  labs(y = "Weight of utterance length")
p1
p2
```

This implementation is somewhat unsatisfactory. It hardcodes variable
names, so it's harder to reuse, and it is complicated to use, requiring
three functions and a join.

In the package implementation, we hide everything behind a single function.

```{r, fig.width = 3, fig.height = 3, out.width = "40%", fig.show = "hold"}
data_plot <- weight_lengths_with_ordinal_model(
  data_train = data_example_intelligibility_by_length, 
  var_length = tocs_level,
  var_x = age_months, 
  id_cols = c(child), 
  spline_df = 3
)
    
p1 <- ggplot(data_plot) + 
  aes(x = age_months, y = tocs_level_prob_reached ) + 
  geom_line(aes(color = ordered(tocs_level))) +
  ggrepel::geom_label_repel(
    aes(label = tocs_level),
    data = data_plot |> 
      filter(age_months == 30), 
    xlim = c(28, 30), 
    seed = 1
  ) +
  ggrepel::geom_label_repel(
    aes(label = tocs_level),
    data = data_plot |> 
      filter(age_months == 47), 
    xlim = c(47, 49), 
    seed = 1
  ) +
  scale_color_ordinal(end = .85) +
  guides(color = "none") + 
  labs(x = "Age [months]", y = "Prob. of reaching length") +
  scale_x_continuous(limits = c(28, 49))
p2 <- p1 + aes(y = tocs_level_weight) + labs(y = "Weight of utterance length")
p1
p2
```




## References

Hustad, K. C., Mahr, T., Natzke, P. E. M., & Rathouz, P. J.
(2020). Development of Speech Intelligibility Between 30 and 47 Months in
Typically Developing Children: A Cross-Sectional Study of Growth. *Journal
of Speech, Language, and Hearing Research*, *63*(6), 1675–1687.
https://doi.org/10.1044/2020_JSLHR-20-00008

Hustad, K. C., Mahr, T., Natzke, P. E. M., & J. Rathouz, P. (2020).
Supplemental Material S1 (Hustad et al., 2020). ASHA journals.
https://doi.org/10.23641/asha.12330956.v1

